\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-07-10 scratch/notes_john/JSS-article/JSS-article-4.Rmd
apply JSS style to JSS-article-4 (first pass)

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly=TRUE)
methods("cv")
```
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

@@

```{r Auto}
data("Auto", package="ISLR2")
summary(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

@@

```{r cv-lm-1}
library("car", quietly=TRUE)

m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
brief(m.auto)

cv(m.auto, confint=TRUE)
```
The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

@@

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations); for example:
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive", confint=TRUE)
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the \code{microbenchmark()} function from the package of the same name  [@Mersmann:2023] for the timings:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE

set.seed(19412) # for reproducibility
print(microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10,
  unit = "relative"
  ), signif = 3)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
```{r polyomial-models}
mlist <- vector(10, mode="list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2] # e.g., the quadratic fit
```
We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
```{r polynomial-regression-CV}
# 10-fold CV
mlist <- models(mlist) # create "modList" object

cv.auto.10 <- cv(mlist, data=Auto, seed=2120)
cv.auto.10[2] # e.g., for quadratic model

# LOO CV
cv.auto.loo <- cv(mlist, data=Auto, k="loo")
cv.auto.loo[2] # e.g., for quadratic model
```
The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}. To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| fig.height = 5,
#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p = 1, 2, \\ldots, 10$.",
#| fig.subcap=c('', ''),
#| fig.ncol = 2, out.width = "50%", fig.align = "center"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

@@

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000]. \code{MathAchieve} comprises individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} contains school-level data:
```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package="nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```
The first few students are in school number 1224 and the last few in school 9586.

@@

Some data-management is required before fitting a mixed-effects model to the HSB data:
```{r HSB-data-management, cache=TRUE}
HSB <- MathAchieve
HSB <- merge(MathAchSchool[, c("School", "Sector")],
             HSB[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))
HSB <- within(HSB, {
  mean.ses <- ave(ses, school)
  cses <- ses - mean.ses
})
```
In the process, we merge variables from the school-level and student-level data sets, and create two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data (summarizing the model with the \code{S()} function from the \pkg{car} package):
```{r HSB-lmer, cache=TRUE}
library("lme4", quietly=TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
S(hsb.lmer, brief=TRUE)
```

We can then cross-validate at the cluster (i.e., school) level,
```{r HSB-lmer-CV-cluster, cache=TRUE}
cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
```
or at the case (i.e., student) level,
```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed=1575)
```
For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: \code{clusterVariables = c("school", "class")}. For cluster-level CV, the default is \code{k = "loo"}, that is, leave one cluster out at a time; we instead specify \code{k = 10} folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

@@

To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^Venables]
```{r Auto-redux}
names(Auto)
xtabs(~ year, data=Auto)
xtabs(~ origin, data=Auto)
xtabs(~ cylinders, data=Auto)
```

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression. (b) Component+residual plots for the working model fit to the \\code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
#| fig.subcap=rep("", 2)
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")

m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```

```{r Auto-explore-show, eval=FALSE}
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```

@@

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model, eval=FALSE}
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```
The component+residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.

@@

The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8] and @FoxWeisberg:2019 [Chap. 3] for details. We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", 
                    "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "zeroth" powers)---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
}
m <- update(m.auto, data=A)
```

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore} (a), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (a)):
```{r Auto-transformed, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the regression model fit to the transformed \\code{Auto} data.",
#| fig.subcap=rep("", 2)
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
crPlots(m, pch=".")
```

```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore} (b), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (b)):
```{r Auto-CR-plots-transformed, eval=FALSE}
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r stepAIC}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
brief(m.step)
```

@@

The \code{"function"} method for \code{cv()} allows us to cross-validate the whole model-selection procedure. The first argument to \code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
```{r Auto-transform-and-select}
num.predictors
cvs <- cv(selectTransStepAIC, data=Auto, seed=76692, 
          working.model=m.auto,
          predictors=num.predictors,
          response="mpg", AIC=FALSE)
cvs
```

@@

Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:
```{r recursive-CV-polynomials}
recursiveCV.auto <- cv(selectModelList, data=Auto, 
                       working.model=mlist,
                       save.model=TRUE, seed=2120)
recursiveCV.auto
brief(recursiveCV.auto$selected.model)
cv(mlist[[7]], seed=2120) # same seed for same folds
```
As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model.

We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE` (identical output not shown):
```{r recursive-cv-alt, eval=FALSE}
cv(mlist, data=Auto, seed=2120, recursive=TRUE, save.model=TRUE)
```

@@

Suppose that we want to cross-validate a multinomial logistic regression model fit by the \code{multinom()} function in the \pkg{nnet} package [@VenablesRipley:2002]. We borrow an example from @Fox:2016 [Sec. 14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election. Data for the example are in the \code{BEPS} data set in the \pkg{carData} package:
```{r BEPS-data}
data("BEPS", package="carData")
summary(BEPS)
```
The polytomous (multi-category) response variable is \code{vote}, a factor with levels \code{"Conservative"}, \code{"Labour"}, and \code{"Liberal Democrat"}. The predictors of \code{vote} are:

@@

The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}, which was the focus of the original research on which this example is based [@AndersenHeathSinnott:2002]; the other predictors enter the model additively:
```{r BEPS-model}
library("nnet", quietly=TRUE)
m.beps <- multinom(vote ~ age + gender + economic.cond.national +
                       economic.cond.household + Blair + Hague + Kennedy +
                       Europe*political.knowledge, data=BEPS, trace=FALSE)

@@

The  interaction is associated with a very small $p$-value. Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study."
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels=list(Europe=1:11, political.knowledge=0:3),
            fixed.predictors=list(given.values=c(gendermale=0.5))),
     lines=list(col=c("blue", "red", "orange")), # party colors
     axes=list(x=list(rug=FALSE), y=list(style="stacked")))
```
As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.

@@

It is straightforward to adapt Bayes's rule to a polytomous response:
```{r BayesRuleMulti}
head(BEPS$vote)
yhat <- predict(m.beps, type="class")
head(yhat)

@@

The marginal proportions for the response categories are
```{r BEPS-response-distribution}
xtabs(~ vote, data=BEPS)/nrow(BEPS)
```
and so the marginal Bayes's rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$. The multinomial-logit model appears to do substantially better than that, but does its performance hold up to cross-validation?

We check first whether the default \code{cv()} method works "out-of-the-box" for the \code{"multinom"} model:
```{r BEPS-test-default, error=TRUE}
cv(m.beps, seed=3465, criterion=BayesRuleMulti)
```
The default method of \code{GetResponse()} (a function supplied by the \pkg{cv} package---see \code{?GetResponse}) fails for a \code{"multinom"} object. A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the `get_response()` function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
```{r GetResponse.multinom}
GetResponse.multinom <- function(model, ...) {
  insight::get_response(model)
}

>>>>

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly = TRUE)
methods("cv")
```
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

@@

```{r Auto}
data("Auto", package = "ISLR2")
summary(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

@@

```{r cv-lm-1}
library("car", quietly = TRUE)

m.auto <- lm(mpg ~ poly(horsepower, 2), data = Auto)
brief(m.auto)

cv(m.auto, confint = TRUE)
```
The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

@@

```{r cv.lm-2`}
cv(m.auto, k = "loo")
```
The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations); for example:
```{r cv.lm-3}
cv(m.auto, k = "loo", method = "naive", confint = TRUE)
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM; the output shown is for the CV estimate of MSE and the bias-corrected CV estimate. We use the \code{microbenchmark()} function from the package of the same name  [@Mersmann:2023] for the timings:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data = Auto)
boot::cv.glm(Auto, m.auto.glm)$delta

set.seed(19412)
print(microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k = "loo"),
  Woodbury = cv(m.auto, k = "loo", method = "Woodbury"),
  naive = cv(m.auto, k = "loo", method = "naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times = 10, unit = "relative"), signif = 3)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
```{r polyomial-models}
mlist <- vector(10, mode = "list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2]
```
where, for example \code{mlist[2]} is the quadratic fit.

We then apply \code{cv()} to the list of 10 models for both 10-fold and LOO CV (the \code{data} argument to \code{cv()} is required):
```{r polynomial-regression-CV}
mlist <- models(mlist)

cv.auto.10 <- cv(mlist, data = Auto, seed = 2120)
cv.auto.10[2]

cv.auto.loo <- cv(mlist, data = Auto, k = "loo")
cv.auto.loo[2]
```
The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}. To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| fig.height = 5,
#| fig.cap = "Cross-validated (a) 10-fold and (b) LOO MSE as a function of polynomial degree, $p = 1, 2, \\ldots, 10$.",
#| fig.subcap=c('', ''),
#| fig.ncol = 2, out.width = "50%", fig.align = "center"
plot(cv.auto.10, main = "Polynomial Regressions, 10-Fold CV",
     axis.args = list(labels = 1:10), xlab = "Degree of Polynomial, p")
plot(cv.auto.loo, main = "Polynomial Regressions, LOO CV",
     axis.args = list(labels = 1:10), xlab = "Degree of Polynomial, p")
```

@@

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000]. \code{MathAchieve} comprises individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} contains school-level data:
```{r HSB-data}
data("MathAchieve", package = "nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package = "nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```
The first few students are in school number 1224 and the last few in school 9586.

@@

Some data-management is required before fitting a mixed-effects model to the HSB data:
```{r HSB-data-management, cache=TRUE}
HSB <- MathAchieve
HSB <- merge(MathAchSchool[, c("School", "Sector")],
             HSB[, c("School", "SES", "MathAch")], by = "School")
names(HSB) <- tolower(names(HSB))
HSB <- within(HSB, {
  mean.ses <- ave(ses, school)
  cses <- ses - mean.ses
})
```
In the process, we merge variables from the school-level and student-level data sets, and create two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data (summarizing the model with the \code{S()} function from the \pkg{car} package):
```{r HSB-lmer, cache=TRUE}
library("lme4", quietly = TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data = HSB)
S(hsb.lmer, brief = TRUE)
```

We can then cross-validate at the cluster (i.e., school) level,
```{r HSB-lmer-CV-cluster, cache=TRUE}
cv(hsb.lmer, k = 10, clusterVariables = "school", seed = 5240)
```
or at the case (i.e., student) level,
```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed = 1575)
```
For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: \code{clusterVariables = c("school", "class")}. For cluster-level CV, the default is \code{k = "loo"}, that is, leave one cluster out at a time; we instead specify \code{k = 10} folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

@@

To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^Venables]
```{r Auto-redux}
names(Auto)
xtabs(~ year, data = Auto)
xtabs(~ origin, data = Auto)
xtabs(~ cylinders, data = Auto)
```

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression. (b) Component + residual plots for the working model fit to the \\code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
#| fig.subcap=rep("", 2)
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight 
                  + acceleration,
                  smooth = list(spread = FALSE), data = Auto, pch= ".")

m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch = ".")
```

```{r Auto-explore-show, eval=FALSE}
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight 
                  + acceleration, 
                  smooth = list(spread = FALSE), data = Auto, pch = ".")
```

@@

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model, eval=FALSE}
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch = ".")
```
The component + residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.

@@

The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8] and @FoxWeisberg:2019 [Chap. 3] for details. We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", 
                    "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "zeroth" powers)---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda = powers[pred])
}
m <- update(m.auto, data = A)
```

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore} (a), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (a)):
```{r Auto-transformed, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component + residual plots for the regression model fit to the transformed \\code{Auto} data.",
#| fig.subcap=rep("", 2)
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread = FALSE), data = A, pch = ".")
crPlots(m, pch = ".")
```

```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth = list(spread = FALSE), data = A, pch = ".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore} (b), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (b)):
```{r Auto-CR-plots-transformed, eval=FALSE}
crPlots(m, pch = ".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r stepAIC}
library("MASS")
m.step <- stepAIC(m, k = log(nrow(A)), trace = FALSE)
brief(m.step)
```

@@

The \code{"function"} method for \code{cv()} allows us to cross-validate the whole model-selection procedure. The first argument to \code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
```{r Auto-transform-and-select}
num.predictors
cvs <- cv(selectTransStepAIC, data = Auto, seed = 76692, 
          working.model = m.auto, predictors = num.predictors,
          response = "mpg", AIC = FALSE)
cvs
```

@@

Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:
```{r recursive-CV-polynomials}
recursiveCV.auto <- cv(selectModelList, data = Auto, 
                       working.model = mlist, save.model = TRUE, 
                       seed = 2120)
recursiveCV.auto
brief(recursiveCV.auto$selected.model)
cv(mlist[[7]], seed = 2120)
```
As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model (using the same seed for \proglang{R}'s random-number generator to obtain the same folds).

We can equivalently call `cv()` with the list of models as its first argument and set the argument `recursive=TRUE` (identical output not shown):
```{r recursive-cv-alt, eval=FALSE}
cv(mlist, data = Auto, seed = 2120, recursive = TRUE, save.model = TRUE)
```

@@

Suppose that we want to cross-validate a multinomial logistic regression model fit by the \code{multinom()} function in the \pkg{nnet} package [@VenablesRipley:2002]. We borrow an example from @Fox:2016 [Sec. 14.2.1], with data from the British Election Panel Study on vote choice in the 2001 British election. Data for the example are in the \code{BEPS} data set in the \pkg{carData} package:
```{r BEPS-data}
data("BEPS", package = "carData")
summary(BEPS)
```
The polytomous (multi-category) response variable is \code{vote}, a factor with levels \code{"Conservative"}, \code{"Labour"}, and \code{"Liberal Democrat"}. The predictors of \code{vote} are:

@@

The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}, which was the focus of the original research on which this example is based [@AndersenHeathSinnott:2002]; the other predictors enter the model additively:
```{r BEPS-model}
library("nnet", quietly = TRUE)
m.beps <- multinom(vote ~ age + gender + economic.cond.national
                        + economic.cond.household + Blair + Hague 
                        + Kennedy + Europe * political.knowledge, 
                   data = BEPS, trace = FALSE)

@@

The  interaction is associated with a very small $p$-value. Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the \code{Europe} $\times$ \code{political.knowledge} interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study, using party colors."
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels = list(Europe = 1:11, political.knowledge = 0:3),
            fixed.predictors = list(given.values = c(gendermale = 0.5))),
     lines = list(col = c("blue", "red", "orange")),
     axes = list(x = list(rug = FALSE), y = list(style = "stacked")))
```
As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was relatively Euro-skeptic, while Labour and the Liberal Democrats were more supportive of the UK's participation in the EU.

@@

It is straightforward to adapt Bayes's rule to a polytomous response:
```{r BayesRuleMulti}
head(BEPS$vote)
yhat <- predict(m.beps, type = "class")
head(yhat)

@@

The marginal proportions for the response categories are
```{r BEPS-response-distribution}
xtabs(~ vote, data = BEPS) / nrow(BEPS)
```
and so the marginal Bayes's rule prediction, that everyone will vote Labour, produces an error rate of $1 - 0.47213 = 0.52787$. The multinomial-logit model appears to do substantially better than that, but does its performance hold up to cross-validation?

We check first whether the default \code{cv()} method works "out-of-the-box" for the \code{"multinom"} model:
```{r BEPS-test-default, error=TRUE}
cv(m.beps, seed = 3465, criterion = BayesRuleMulti)
```
The default method of \code{GetResponse()} (a function supplied by the \pkg{cv} package---see \code{?GetResponse}) fails for a \code{"multinom"} object. A straightforward solution is to supply a \code{GetResponse.multinom()} method that returns the factor response [using the `get_response()` function from the \pkg{insight} package, @LudeckeWaggonerMakowski:2019],
```{r GetResponse.multinom}
GetResponse.multinom <- function(model, ...) {
  insight::get_response(model)
}
